# Reproducible Research: Peer Assessment 1

This is the Rmd file of the Peer Assessment No. 1 for Reproducible Research in Coursera.

The primary objectives of this assignment are:
 1. To read the `activity.csv` file from working directory
 2. To determine total steps taken each day, draw a histogram, and determine mean and median of daily steps.
 3. To draw a time series plot of average steps taken in 5 minute interval across all days, and to determine the 5 minute interval that contains maximum average steps taken.
 4. To determine the missing values and impute the missing values.  Compare the dateset that contains imputed values to the original dataset.
 5. To compare the activities between weekday and weekend.
 
The data was downloaded from the assignment website on June 5, 2014 and unzipped using the following codes:

`fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"`

`download.file(fileURL, dest="activity.zip", method = "auto")`

`unzip("activity.zip")`

I avoid running the code repeatedly to prevent overload of the website.
 
## Loading and preprocessing the data

The following codes read the `activity.csv` file.  Transformation is not necessary at this time to perform the required tasks.
```{r, echo=TRUE}
activity <- read.csv("activity.csv", header = T)
summary(activity)
```
## What is mean total number of steps taken per day?

Total activity for each day, mean and median of total activity were calculated.
```{r, echo=TRUE}
total.activity <- aggregate(activity$steps, list(activity$date), sum)
names(total.activity) <- c("date","steps")
head(total.activity)
dailymean <- round(mean(total.activity$steps, na.rm=T), digits=2)
dailymedian <- median(total.activity$steps, na.rm=T)
print(paste("Mean of the total daily activity:", dailymean, "steps."))
print(paste("Median of the total daily activity:", dailymedian, "steps."))
```
Draw histogram with `ggplot2`
```{r fig1, echo=TRUE}
library(ggplot2)
p1 <- ggplot(total.activity, aes(steps)) + geom_histogram(fill="blue", color="yellow", binwidth = 1000)
q1 <- p1 + ggtitle("Figure 1. Histogram of Daily Total Steps\nfrom 2012-10-01 to 2012-11-30") +   xlab("Daily Total Steps") + theme(plot.title = element_text(face="bold")) + 
     geom_vline(xintercept=dailymean, color = "red", alpha = 0.4, size = 1.2) +
     geom_vline(xintercept=dailymedian, color = "green", alpha = 0.4, size = 1.2) +
     annotate("text", x = 13600, y = 10.3, label = "mean/median", color = "orange")
print(q1)
```
In Figure 1, the red vertical line is mean, and the green vertical line is median.  Because these two values are very close, the lines are superimposed.

## What is the average daily activity pattern?

Average steps taken in each time interval is calculated.  The interval is not continuous.  They represent hour and minute of military time.  There is no interval between 60 and 95.  For example, interval 2005 represent 20:05.
```{r, echo=TRUE}
timed.activity <- aggregate(activity$steps, list(activity$interval), mean, na.rm = TRUE)
names(timed.activity) <- c("interval", "steps")
head(timed.activity)
timed.activity$interval
```
Finding in which 5-minute interval lies maximal average steps:
```{r, echo=TRUE}
a <- timed.activity[timed.activity$steps==max(timed.activity$steps),]
print(paste("The maximaum of", round(a$steps, digits=1), "steps is located in the interval", a$interval))
```
The interval is **835** (8:35).

The data is plotted with `ggplot2`.
```{r fig2, echo=TRUE}
p2 <- ggplot(activity, aes(interval, steps)) + stat_summary(fun.y = mean, geom = "line", color = "blue", na.rm = TRUE)
q2 <- p2 + ggtitle("Figure 2. Time Series Plot of Average Steps in Each Interval") +
     theme(plot.title = element_text(face = "bold")) + ylab("Average Number of Steps") +
     xlab("Interval")
print(q2)
```

## Imputing missing values

Finding where the `NA`s are:
```{r, echo=TRUE}
table(activity$date, is.na(activity$steps))
```

The all-day data from the dates 2012-10-01, 2012-10-08, 2012-11-01, 2012-11-04, 2012-11-09, 2012-11-10, 2012-11-14, and 2012-11-30 are `NA`s.  There are a total of **2304** `NA`s.

To impute missing values, I created a data frame of mean and standard deviation for each interval.  The missing values were imputed by random number generated from mean and standard deviation for the interval.
```{r, echo=TRUE}
sd.activity <- tapply(activity$steps, as.factor(activity$interval), sd, na.rm=TRUE)
meansd <- cbind(timed.activity, sd.activity)
names(meansd) <- c("interval", "mean", "sd")
```
The following function will take in the interval of the missing value, convert the interval to row number, and generate random number with reference to mean and standard deviation from the `meansd` table.  Because the steps cannot be negative, if the generated random number is a negative number, the value is assigned as 0.
```{r, echo=TRUE}
imputeNA <- function(interval) {
     a <- interval %/% 100
     b <- interval %% 100
     rownum <- a*12 + b/5 + 1
     c <- round(rnorm(1, mean = meansd[rownum, 2], sd = meansd[rownum, 3]))
     if (c < 0) {c <- 0}
     return(c)
}
```
The following codes will create dataset `activity2` from `activity`.  If the data is missing, it will create the data using `imputeNA` function.  If the data is not missing, it will simply copy the data from `activity`.
```{r, echo=TRUE}
activity2 <- data.frame(steps=numeric(), date=as.Date(character()),
                        interval=numeric())
n <- nrow(activity)
set.seed(123456)
for (i in 1:n) {
     if (is.na(activity[i,1])) {
          activity2[i,1] <- imputeNA(activity[i,3])
          activity2[i,2:3] <- activity[i,2:3]
     }
     else {activity2[i,1:3] <- activity[i,1:3]}
}
summary(activity2)
```
Now the dataset `activity2` no longer has `NA`s.  The dataset is saved as `activity2.csv` in the working directory.
```{r, echo=TRUE}
write.csv(activity2, "activity2.csv", row.names = FALSE)
```
Average daily steps are recalculated using `activity2`.  The new mean and median values are calculated.
```{r, echo=TRUE}
total.activity2 <- aggregate(activity2$steps, list(activity2$date), sum)
names(total.activity2) <- c("date","steps")
dailymean2 <- round(mean(total.activity2$steps), digits=2)
dailymedian2 <- median(total.activity2$steps)
print(paste("Mean of the new total daily activity:", dailymean2, "steps."))
print(paste("Median of the new total daily activity:", dailymedian2, "steps."))
```
As a result of imputing the missing values, both mean and median of daily total steps slightly increase, but the 2 values are still very close.  The median is slightly higher than mean.  It appeared that the imputation assigned around 15000 daily steps in the dates with missing values.  As a result, there is a tall peak around 15000 steps, which shifted the mean and median.  The following is the histogram of daily total steps using the new dataset.
```{r fig3, echo=TRUE}
p3 <- ggplot(total.activity2, aes(steps)) + geom_histogram(fill="blue", color="yellow", binwidth = 1000)
q3 <- p3 + ggtitle("Figure 3. Histogram of Daily Total Steps\nafter Imputing Missing Values") +   xlab("Daily Total Steps") + theme(plot.title = element_text(face="bold")) + 
     geom_vline(xintercept=dailymean2, color = "red", alpha = 0.4, size = 1.2) +
     geom_vline(xintercept=dailymedian2, color = "green", alpha = 0.4, size = 1.2) +
     annotate("text", x = c(10200,13000), y = 10.3, label = c("mean","median"), color = c("red","green"))
print(q3)
```

## Are there differences in activity patterns between weekdays and weekends?

Adding a column `wd` to separate weekday from weekend.  I use `weekdays()` to find the day of the week.  If the result is "Sat" or "Sun", weekend is assigned.  Otherwise, weekday is assigned.
```{r, echo=TRUE}
activity2$wd <- ifelse(weekdays(activity2$date, abbreviate = TRUE) %in% c("Sat","Sun"),"weekend","weekday")
```
The time series plot of average steps is drawn using `ggplot2` with `weekday` and `weekend` separated out by `facet_grid`.
```{r fig4, echo=TRUE}
p4 <- ggplot(activity2, aes(interval, steps)) 
q4 <- p4 + stat_summary(fun.y = mean, geom = "line", color = "blue") +
     facet_grid(wd ~.) + ggtitle("Figure 4. Time Series Plot Comparing Weekdays and Weekends") +
     theme(plot.title = element_text(face = "bold")) + xlab("Interval") +
     ylab("Average Number of Steps")
print(q4)
```
The plot showed that in the weekdays, there is a peak around interval 835 with increased activity from interval 500.  In the weekends, the activity was spreaded out 750 to 2100.


